---
title: "CSP-571 Project - Predicting Mushroom Edibility"
author: "Liwei He"
output: html_document
---

```{r setup, include=FALSE}
required_packages <- c(
  "knitr",
  "ggplot2",
  "scales",
  "rpart",
  "rpart.plot",
  "randomForest",
  "pROC"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    paste(
      "Install these packages before knitting:",
      paste(missing_packages, collapse = ", ")
    )
  )
}

project_root <- normalizePath("..", winslash = "/", mustWork = TRUE)
knitr::opts_knit$set(root.dir = project_root)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "90%"
)
```

# 1. Project Goal

Link to the dataset:

Biology  
Mushrooms: https://visualization.group/data/mushroom/

I'm using the
[UCI Secondary Mushroom Dataset](https://archive.ics.uci.edu/dataset/848/secondary+mushroom+dataset)
to predict whether a simulated mushroom record is edible or poisonous. 

I built and compared three models: logistic regression, a decision tree, and a random forest.

# 2. Data Reading and Cleaning

The raw file is semicolon-delimited and contains one target variable and 20 predictors.

```{r load-data}
RUN_QC <- FALSE
RUN_PLOTS <- FALSE
source("code/01_eda.R")

data_summary <- data.frame(
  Item = c(
    "Raw rows",
    "Columns",
    "Numerical predictors",
    "Categorical predictors"
  ),
  Value = c(
    nrow(mushrooms),
    ncol(mushrooms),
    length(numeric_variables),
    length(categorical_variables)
  )
)

knitr::kable(data_summary, caption = "Data overview")
```

9 categorical predictors have missing values, some quite heavily. 
Rather than dropping those rows, I recoded missing entries as an explicit `unknown`
level, and removed 146 exact duplicate rows before splitting the data.

```{r cleaning-summary}
knitr::kable(
  cleaning_summary,
  col.names = c("Cleaning check", "Value"),
  caption = "Cleaning results"
)
```

The final modeling dataset contains 60,923 rows, no missing values, and no exact
duplicates.

# 3. Exploratory Data Analysis

## Class distribution

```{r class-plot, fig.height=5}
eda_plots$class_distribution
```

The classes are close to balanced - 44.6% edible, 55.4% poisonous - so
accuracy is a reasonable metric here. I also track poisonous-class recall,
since calling a poisonous mushroom "edible" is the costlier mistake.

## Missing values

```{r missing-plot, fig.height=6}
eda_plots$missing_values
```

A handful of variables are missing more than 80% of their values, which ruled
out dropping incomplete rows. I kept them and coded missing entries as their
own `unknown` level - if missingness itself correlates with edibility, the
model can use that signal instead of losing it.

## Numerical predictors

```{r numeric-plot, fig.width=11, fig.height=5}
eda_plots$numeric_boxplots
```

Edible mushrooms tend to run a bit larger - bigger cap diameter, taller and
wider stems, though all three measurements have long right tails. I left
those outliers in; they're part of the simulated data as generated.

## Selected categorical predictors

```{r categorical-plot, fig.height=10}
eda_plots$categorical_class_proportions
```

I selected `cap_shape`, `gill_attachment`, and `habitat` for this plot -
together they cover cap morphology, gill structure, and growing environment,
and their levels show visibly different class proportions. A few habitat
categories are entirely one class, but those are also the smallest groups.
Most common levels still contain a mix of both classes, which is the point of
combining several predictors in a model.

# 4. Train/Test Split

I use `set.seed(571)` and an 80/20 stratified split. All three models are
evaluated on the same held-out test set.

```{r run-models, results='hide'}
RUN_QC <- FALSE
RUN_PLOTS <- FALSE
RUN_MODEL_PLOTS <- FALSE
source("code/02_modeling.R")
```

```{r split-table}
knitr::kable(split_summary, caption = "Stratified data split")
```

The class proportions are nearly identical across the training and test sets.

# 5. Logistic Regression

The logistic model provides a linear baseline. Numerical predictors are
standardized using training-set statistics only.

```{r logistic-results}
knitr::kable(
  as.data.frame.matrix(logistic_results$confusion_matrix),
  caption = "Logistic regression confusion matrix"
)
```

Logistic regression is the weakest of the three, with 848 dangerous false
negatives. R also warns that some fitted probabilities are numerically 0 or
1, a sign the classes are close to perfectly separable on some combination of
predictors - logistic regression just isn't flexible enough to find that
boundary. I use it mainly as a baseline rather than digging into individual
coefficients.

# 6. Decision Tree

```{r tree-results}
knitr::kable(
  as.data.frame.matrix(tree_results$confusion_matrix),
  caption = "Decision tree confusion matrix"
)
```

The full tree does much better - 98.91% test accuracy, 60 dangerous false
negatives. It's too large to plot legibly, so the figure below is a pruned
version showing the top few splits; the reported numbers still come from the
full tree.

```{r tree-plot, fig.width=11, fig.height=7}
display_candidates <- tree_model$cptable[
  tree_model$cptable[, "nsplit"] <= 6,
  ,
  drop = FALSE
]

display_cp <- display_candidates[nrow(display_candidates), "CP"]
display_tree <- rpart::prune(tree_model, cp = display_cp)

rpart.plot::rpart.plot(
  display_tree,
  type = 2,
  extra = 104,
  fallen.leaves = TRUE,
  tweak = 0.9,
  main = "Simplified Decision Tree"
)
```

Stem width is the first split, then cap color, then stem root. It's a clean
illustration of how a tree turns classification into readable if/then rules
- though these specific thresholds are fit to one simulated dataset, not
mycological fact.

# 7. Random Forest

The random forest combines 300 trees and provides an out-of-bag error estimate.

```{r forest-results}
knitr::kable(
  as.data.frame.matrix(random_forest_results$confusion_matrix),
  caption = "Random forest confusion matrix"
)

oob_result <- data.frame(
  Measure = "OOB error",
  Value = round(as.numeric(random_forest_oob_error), 6)
)

knitr::kable(oob_result, caption = "Random forest OOB result")
```

The random forest gets every test record right, with zero OOB error. This
matches the dataset's original paper, so it's not a bug on my end - the
simulated data is just that separable. It's a strong result on this
benchmark, not proof the model would generalize to a real, unseen species.

```{r importance-plot, fig.width=9, fig.height=7}
randomForest::varImpPlot(
  random_forest_model,
  type = 1,
  n.var = 15,
  main = "Random Forest Variable Importance"
)
```

```{r numeric-correlation-check}
correlation_check <- data.frame(
  Comparison = c(
    "stem_width vs. cap_diameter",
    "stem_width vs. stem_height"
  ),
  Pearson = c(
    cor(mushrooms_clean$stem_width, mushrooms_clean$cap_diameter, method = "pearson"),
    cor(mushrooms_clean$stem_width, mushrooms_clean$stem_height, method = "pearson")
  ),
  Spearman = c(
    cor(mushrooms_clean$stem_width, mushrooms_clean$cap_diameter, method = "spearman"),
    cor(mushrooms_clean$stem_width, mushrooms_clean$stem_height, method = "spearman")
  )
)

knitr::kable(
  correlation_check,
  digits = 3,
  caption = "Correlation check for numerical size predictors"
)
```

`cap_surface`, `gill_attachment`, `cap_color`, and `gill_color` top the
random forest's permutation-importance ranking. `stem_width` - the first
split in the simplified tree - ranks around 11th. That's not a contradiction:
a single tree picks the locally best one-step split, while permutation
importance measures the accuracy drop from shuffling a variable across the
whole forest. The correlation check shows why: `stem_width` overlaps heavily
with `cap_diameter` (r = 0.695, Spearman = 0.863), more than with
`stem_height`. Once `cap_diameter` is available, shuffling `stem_width`
doesn't cost the forest much - it can recover similar information elsewhere.
The ranking reflects each variable's marginal contribution within this
forest, not a causal claim about what makes a mushroom poisonous.

# 8. Model Comparison and Conclusion

```{r comparison-table}
comparison_table <- model_comparison

percentage_columns <- c(
  "Accuracy",
  "Poisonous_Recall",
  "Specificity",
  "Precision",
  "F1"
)

comparison_table[percentage_columns] <- lapply(
  comparison_table[percentage_columns],
  scales::percent,
  accuracy = 0.01
)

knitr::kable(
  comparison_table,
  caption = "Held-out test-set model comparison"
)
```

Random forest wins outright - perfect metrics, zero dangerous false
negatives. The decision tree is close behind at 98.91% accuracy and has one
real advantage: you can explain the rule to someone. Logistic regression
trails both by a wide margin, which suggests the edible/poisonous boundary
here isn't linear. 
There's enough interaction between stem dimensions, color and surface texture, that a straight-line model can't capture it.

I will go with the random forest: the cost of a false negative is too high to
trade away for interpretability. The decision tree as the easy-to-explain version, since a 1-2% accuracy gap is okay for something you can actually walk someone through.

One caveat on the perfect random forest score: all 61,069 records come from
the same simulation process built on only 173 real species, so a random
train/test split isn't really testing on an unseen species - it's closer to
testing on a held-out sample from the same distribution. A species-aware
split would be a tougher, more realistic test. I didn't have time to build
that here, but it's the first thing I'd try next, along with checking how
much accuracy holds up using just the top 3-4 features.

```{r session-info}
sessionInfo()
```

---

**AI Assistance Note**

The explanation of why `stem_width` was the top split in the pruned decision
tree but ranked around 11th in the random forest's permutation importance
came out of a discussion with Claude (Sonnet 5). 
